---
title: "BaMM analysis"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, warning=FALSE, echo=T, message=FALSE}
library(dgRaph)
library(dplyr)
```

Using the [BaMM software](https://github.com/soedinglab/BaMMmotif) we obtain the transition probabilities for a second order Markov chain motif model for the CTCF motif. We then encode the second order Markov chain as a first order Markov chain by compounding variables $X_1$, $(X_1, X_2)$, $(X_2, X_3)$, $\ldots$, $(X_{15},X_{16})$. We provide the transition probabilites for the Markov chain as a list of matrices, `BammTransProb`. The first matrix is a 1x4 matrix giving the distribution of $X_1$, the second is a 4x16 matrix giving the conditional distribution of $(X_1,X_2$ given $X_1$, the following matrices are 16x16 matrices giving the conditional distribution of $(X_{i+1},X_{i+2})$ given $(X_i,X_{i+1})$. 

```{r}
BammTransProb[[1]]
BammTransProb[[2]]
```

We also provide the background transition probabilities `BammTransProbBg`. We can now build the foreground and background models in the `dgRaph`. 

```{r buildModels}
varDim <- c(4, rep(16, 16-1))
facNbs <- c(list(1),lapply(1:15, function(i) c(i, i+1) ))
bamm_dfg <- dfg(varDim, BammTransProb, facNbs)

potMap <- c(1,2,rep(3, 16-2))
bamm_dfg_bg <- dfg(varDim, BammTransProbBg, facNbs, potMap = potMap)
```

Next we evaluate the significance of a range of scores using saddlepoint approximation and importance sampling.

```{r evalSignificance}
score_dist_saddle <- tailSaddle(seq(-15, 5, 0.05), bamm_dfg_bg, bamm_dfg)
score_dist_saddle$method <- "SA"
score_dist_is <- tailIS(seq(-15, 5, 0.05), 
                        n = 1000, 
                        alpha = c(0.1,0.2,0.3,0.4,0.5), 
                        dfg1 = bamm_dfg_bg, dfg2 = bamm_dfg) %>%
  as.data.frame()
score_dist_is$alpha <- as.factor(score_dist_is$alpha)
score_dist_is$method <- "IS"
```

```{r tailPlot, warning=FALSE, message=FALSE}
library(ggplot2)

plot_df <- rbind(score_dist_saddle[,c("x","p","method")],
                 score_dist_is[,c("x","p","method")])

ggplot(plot_df, aes(x = x, y = p, colour = method)) + 
  geom_line() +  scale_y_log10() + theme_bw() + 
  geom_ribbon(data = score_dist_is, aes(ymax = high, ymin = low, fill = alpha, group = alpha), alpha = 0.3)
```

By using different alphas for different parts of the tail. The confidence intervals around the estimates has uniform width on log scale, corresponding to uniform relative error. 