<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{An R Markdown Vignette with knitr}
-->
```{r setup, include=FALSE}
library(knitr)
library(PGMscore)
require(igraph)
library(ggplot2)
```

## PGMscore tutorial
A factorgraph consists of variables and factors. 

```{r echo=TRUE, warning=FALSE}
varDim <- rep(4,4)
facPot <- c(list(matrix(0.25,1,4)),
            list(matrix(0.25,4,4)),
            list(matrix(0.25,4,4)),
            list(matrix(0.25,4,4)))
facNbs <- c(list(c(1L)),
            list(c(1L,2L)),
            list(c(1L,3L)),
            list(c(3L,4L)))
pgm <- PGM(varDim, facPot, facNbs)
```

If you have `igraph` installed the structure of the graph can be plotted using `plot`
```{r}
plot(pgm)
```

#### Tail approximation
Suppose we have two models with the same graph structure, but different factor potentials. We can define a score of an observation as the loglikelihoodratio
$$
S(x) = \log \frac{\prod_{\mathcal{A}}f_a^\prime(x_a)}{\prod_{\mathcal{A}}f_a(x_a)} = \sum_{\mathcal{A}} \left[\log f_a^\prime(x_a) - \log f_a(x_a)\right]
$$

Bayesian statistics...

Lets define a foreground model where the variables are highly correlated with the ancestors
```{r}
set.seed(1)
facPotFg <- c(list(rndDist(1,4)),
              list(rndDist(4,4)),
              list(rndDist(4,4)),
              list(rndDist(4,4)))
```

Lets calculate the tail probabilities in different points using the saddlepoint approximation
```{r}
saddlePointTails(3, pgm, facPotFg)
x <- seq(1,3,by=0.1)
cdf_upper_tail <- sapply(x,FUN=saddlePointTails, pgm=pgm, facPotFg=facPotFg)
df_tail <- data.frame(x = x, tail = cdf_upper_tail)

library(ggplot2)
ggplot(df_tail, aes(x=x,y=tail)) + geom_line() +
  theme_bw() + 
  ylab("p-value") + 
  xlab("score") +
  ggtitle("Saddlepoint approximation of the upper tail")
```

Simulation of tail
```{r}
sampleIS <- sampleISCpp(1000, 0.0, varDim, facPot, facNbs, facPotFg)

x <- seq(1,3,by=0.1)
cdf_upper_tail <- sapply(x, function(x){sum(sampleIS$weights[sampleIS$scores>x])/1000})
df_tail <- data.frame(x = x, tail = cdf_upper_tail)

library(ggplot2)
ggplot(df_tail, aes(x=x,y=tail)) + geom_line() +
  theme_bw() + 
  ylab("p-value") + 
  xlab("score") +
  ggtitle("Importance sampling of the upper tail")
```


